---
title: "The TRANSFORM Trial: Sample Size Justification and Power Analysis"
format:
  html:
    embed-resources: true
editor: visual
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false
library(pwr)
library(tidyverse)
```

### Sample size justification

This study is a prospective, randomised, double-blind Phase II pilot clinical trial comparing MDMA psychedelic assisted psychotherapy (MDMA-PAP) with an active comparator (dexamfetamine) in the treatment of co-morbid chronic pain and post-traumatic stress disorder (PTSD).

The primary endpoint is the mean change in pain intensity — assessed on a 0 to 10 numerical rating scale — between MDMA-PAP (treatment) and dexamfetamine (control) groups, measured at 1 month post treatment completion.

This sample size calculation is based on data from [@christie2022]. In this study, the Chronic Pain Grade (CPG) mean values were 67 pre-treatment and 51.9 post-treatment, with a p-value of 0.034 in the hight pain intensity group. The estimated standard deviation (SD) from this data was calculated as follows:

```{r}
# Sample size of the high-intensity group
n <- 9

# Pre- and post-treatment mean values
pretxt <- 67
postxt <- 51.9

# p-value
p_value <- 0.034

# Degrees of freedom
df <- n - 1

# t-statistic for a one-sample test
t_stat <- qt(1 - p_value/2, df)

# Sample standard deviation calculation
sd <- abs(pretxt - postxt) / t_stat * sqrt(n)

cat('Standard deviation:', round(sd))
```

Then, assuming a within-group standard deviation of 18, a mean change of 15 points in pain intensity between participants treated with MDMA-PAP and dexamfetamine (with no change expected in the control group over time), the power for detecting this difference with a sample size of 20 participants per group is calculated as follows:

```{r}
power.t.test(delta = 15, sd = round(sd), n=20)
```

Thus, a total sample size of 40 participants (20 per group) will provide more than 70% power to detect a statistically significant difference in the mean change from baseline in pain intensity scores at 1 month post treatment completion.

### Other Studies

A placebo-controlled Phase 3 trial published by [@mitchell2023] evaluated the effects of MDMA-assisted therapy compared to placebo in participants with moderate to severe PTSD. The mean change in total severity score from baseline to 18 weeks was -23.7 for MDMA-AT compared to -14.8 for placebo with therapy, resulting in a treatment difference of -8.9 (-13.70, -4.12), with a p-value \< 0.001. The effect size (Cohen’s d) for MDMA-AT versus placebo with therapy was calculated as d = 0.7.

From this study, the standard deviation of the difference between treatment and control was:

```{r}
mitchell_sd <- 8.9/0.7
mitchell_sd
```

Then, with 70% power, the sample size is estimated as:

```{r}
power.t.test(delta=8.9, sd = round(mitchell_sd), power = 0.7)
```

This result align closely with our previous calculations, providing additional support for the proposed sample size.
